---
title: "Interacting with APIs: Example with Spoonacular Food and Recipe API"
author: "Rina Deka"
date: "06/25/2023"
output: 
  github_document:
    toc: true
    html_preview: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


This document is a vignette to demonstrate how one may retrieve data from an [API](https://www.redhat.com/en/topics/api/what-are-application-programming-interfaces). In this tutorial, I will be interacting with Spoonacular's [food API](https://spoonacular.com/food-api/docs). I have created a few functions to interact with a few endpoints for user ease, and I'll use these to explore some of the data that was retrieved. 


# Requirements 

To use the functions for interacting with the food API, I used the following packages:

- [`httr`](https://cran.r-project.org/web/packages/httr/index.html) : 
- [`jsonlite`](https://cran.r-project.org/web/packages/jsonlite/): API interaction
- [`tidyverse`](https://www.tidyverse.org/): tons of useful features for data manipulation and visualization

Please note that the API key I used was "d3ac9a84da954e469bb394dc8b157681", but yours will be different.


```{r,echo=FALSE,warning=FALSE,message=FALSE}
library(knitr)
library(rmarkdown)
library(httr)
library(jsonlite)
library(tidyverse)
```

j
# API Interaction Functions

Suppose that I would like to create a function that will return well-parsed data so that one can look at ingredient and nutrition information for a reciple given some sort of dietary restriction. 

```{r}
getRecipesByDiet <- function(diet,number=10) {
  apiKey <- "d3ac9a84da954e469bb394dc8b157681"  # Replace with your actual API key
  
  endpoint <- "/recipes/complexSearch"
  
  # Set the parameters for the API request
  parameters <- list(
    apiKey = apiKey,
    number = number,  # Number of recipes to retrieve (default: 10)
    diet = diet,  # Dietary restriction (e.g., "vegan", "gluten-free", "keto")
    addRecipeNutrition = TRUE  # Include additional recipe information
    # You can add more parameters as required by the endpoint
  )
  
  # Build the URL
  base_url <- "https://api.spoonacular.com"
  url <- paste0(base_url, endpoint)
  
  # Send the API request
  response <- GET(url, query = parameters)
  
  # Check for successful response
  if (http_type(response) == "application/json") {
    content <- fromJSON(rawToChar(response$content), flatten = TRUE)
    
    # Extract relevant information from the response
    recipes <- content$results
    recipes <- recipes %>% select(-c("gaps","sourceUrl","imageType","image","analyzedInstructions","license","spoonacularSourceUrl","creditsText","author"))
   # recipes <- unnest(recipes,cols=c("nutrition.ingredients"),names_repair = "unique")
    #recipes <- unnest(recipes,cols=c("nutrition.nutrients"),names_repair = "unique")
    #reciples<- recipes %>% unnest(cols=c("nutrition.nutrients","nutrition.properties","nutrition.flavonoids","nutrition.ingredients"),names_repair="unique")
    #Unnest the data-frame and create new columns for each nutrient
    recipes <- unnest_wider(recipes,)

    return(recipes)
  } else {
    stop("API request failed. Status code: ", status_code(response))
  }
}
```



# Exploratory Data Analysis 


```{r}
diet <- "vegan"  # Replace with the desired dietary restriction

recipes <- getRecipesByDiet(diet,number=100)

recipes
```


```{r}
colnames(recipes)
```

```{r}
recipes$nutrition.nutrients
```


```{r}
recipes$nutrition.ingredients
```

```{r}
recipes$nutrition.caloricBreakdown.percentCarbs
```


# Conclusion 





